TTS2016R: A dataset to study population and employment patterns from the 2016 Transportation Tomorrow Survey (TTS) in the Greater Toronto and Hamilton Area, Canada

### 2022-10-23 Manuscript-Data-Package/Manuscript-Data-Package.tex
Edits to draft manuscript - more to be done.

\begin{abstract}
This paper describes and visualises the data contained within the
\{TTS2016R\} data-package created in \texttt{R}, the statistical
computing and graphics language. In addition to a synthetic example,
\{TTS2016R\} contains home-to-work commute information for the Greater
Golden Horseshoe area in Canada retrieved from the 2016 Transportation
Tomorrow Survey (TTS). Included are all Traffic Analysis Zones (TAZ),
the number of people who are employed full-time per TAZ, the number of
jobs per TAZ, origin-destination trips, and calculated car travel time
from TAZ origin-destination centroid pairs. To illustrate how this
information can be analysed to understand patterns in commuting, we
estimate a distance-decay curve (i.e., impedance function) for the
region. \{TTS2016R\} can be freely downloaded and explored at:
\url{https://github.com/soukhova/TTS2016R} where the documentation and
code involved in data creation, manipulation, and the final data
products are detailed.
\end{abstract}

\keywords{Jobs; population; travel time; impedance; Greater Toronto and
Hamilton Area; Ontario, Canada; R}

@@

This manuscript presents the open data product
\href{https://github.com/soukhova/TTS2016R}{\{TTS2016R\}}. Open data
products are the result of turning source data (open or otherwise) into
accessible information that adds value to the original inputs
\citep[see][]{Arribas2021open}. The product presented in this paper is a
\texttt{R} data package which currently consists of a fusion of objects
from a variety of sources: home-to-work flows and trips by mode are
sourced from the 2016 Transportation Tomorrow Survey (TTS)
\citep{data_management_group_tts_2018}, estimated travel times are
calculated, and boundary files from the TTS and from the Canadian
Census.

What is a \texttt{R} data package? A \texttt{R} data package contains
code, data, and documentation in a standardised collection format that
can be installed by \texttt{R} users through a centralized software
repository such as CRAN (the Comprehensive R Archive Network) and
GitHub. \{TTS2016R\} is freely available on GitHub for all to install
and freely use in the spirit of open and reproducible research.
Currently, \{TTS2016R\} includes full-time home-based work-to-job origin
destinations counts and mode-specific trip numbers retrieved from the
2016 TTS, traffic analysis zone (TAZ) boundaries, and municipality,
planning, and census metropolitan area boundaries for the Greater Golden
Horse area (GGH) located in southern Ontario, Canada. In addition, the
package includes TAZ centroid-to-centroid travel times by car, transit,
cycling, and walking mode computed using package \{r5r\}
\citep{Pereira2021r5r}.

The aim of this paper is to walk readers through the data sets,
illustrate a use case (i.e., the calculation of an impedance function
that can be used to calculate accessibility to employment), and invite
others to experiment in its uses and applications. Though data from the
TTS is freely available to the public through the
\href{https://dmg.utoronto.ca/idrs/index}{TTS Data Retrieval System},
the raw data can be technically demanding, cumbersome to work with, and
requires multiple software to process. By pre-processing the data in a
\texttt{R} environment and packaging it alongside complimentary data,
\{TTS2016R\} offers a slice of the TTS data which can be immediately
used by \texttt{R} users to analysis patterns of commuting to work in
the region. Anticipate this package to grow in the future: it currently
provides an open infrastructure for additional TTS or complimentary data
sets to be amended by the authors and the open-source community in the
future by request.

@@

This data is aggregated and available at the level of TAZ: TAZ are a
spatial unit of analysis typically used to estimate the number of trips
produced and attracted to each zone \citep{meyer_urban_2001}. They are
thus defined by transportation planners for a region based on
intra-similarity and inter-dissimilarity between land-use and population
demographics. Within the GGH boundaries, 3,764 TAZ are specified and
each TAZ is uniquely identified using the GTA06 Zoning System: the
survey boundary is discussed in the 2016 TTS methodology and defined by
the TTS \citep{data_management_group_tts_2018}. The TAZ range between
\(\ge\)\texttt{0.019} km\^{}2 in spatial area to a maximum of 879
km\^{}2 (median: 1.3 km\^{}2 and 3rd quantile: 2.8 km\^{}2).

@@

\caption{\label{fig:TTS-16-survey-area}TTS 2016 study area within the GGH in Ontario, Canada along with associated descriptive statistic of workers and jobs per TAZ, OD links (count of workers potentially interacting with their place of employment) by origin TAZ, OD trips by origin TAZ, and calculated OD car travel time (TT) per origin TAZ. 3,507 trips were not assigned travel times as they are longer than 180 mins. Spatial boundary files are retrieved from the TTS which define the survey area: the 20 regions in the GGH are represented by black lines and labelled, the dark gray lines are planning boundaries.}\label{fig:TTS-16-survey-area}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/tts-workers-jobs-plot-1} \caption{\label{fig:tts-workers-jobs-plot}Number of workers (left) and jobs (right) in each TAZ in the 2016 TTS.}\label{fig:tts-workers-jobs-plot}
\end{figure}

Figure \ref{fig:tts-workers-jobs-plot} presents the number of employed
people and associated jobs per TAZ. It can be observed that the spatial
distribution of jobs and workers is unequal, which is indicative of a
jobs -housing imbalance that can impact accessibility in a region
\citep{Levine1998rethinking}. It can also be seen that there is a higher
number of TAZ with no workers than zones with no jobs (i.e., 791 TAZ
with no workers : 396 TAZ with no jobs) and the mean of workers per TAZ
is higher than the mean of jobs (i.e., 916 workers : 819 jobs) the
number TAZ with an extreme number of jobs at the highest and lowest
percentiles is significantly higher than the number of workers.

@@

Also included in \{TTS2016R\} are the estimated travel times between OD
as summarized in descriptive statistics table in Figure
\ref{fig:plot-tt-ttpertrip}; travel times are calculated using the
package \{r5r\}. \{r5r\} interfaces with the java-based \{r5\} routing
engine developed separately by \href{https://conveyal.com/}{Conveyal}.
The inputs to \{r5r\} for this data package were the desired mode, a
maximum travel time threshold of 180 minutes, the geo-coded origin
destination pairs based on the centroids of the TAZ, and the static Open
Street Map road network of Ontario. A travel time threshold of 180
minutes was selected since it captures almost all potential OD
interactions.

@@

\begin{figure}
\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/plot-tt-ttpertrip-1} \caption{\label{fig:plot-tt-ttpertrip}Calculated total worker travel time (left) and average worker travel time (right) for each TAZ in the 2016 TTS. Planning boundaries of Niagara, Waterloo, and the Toronto census metropolitan area are drawn with purple, brown and blue borders, respectively, and discussed. }\label{fig:plot-tt-ttpertrip}
\end{figure}
\newpage

>>>>

\begin{abstract}
This paper describes and visualises the data contained within the
\{TTS2016R\} data package created in \texttt{R}, the statistical
computing and graphics language. In addition to a synthetic example,
\{TTS2016R\} contains home-to-work commute information for the Greater
Golden Horseshoe (GGH) area in Canada retrieved from the 2016
Transportation Tomorrow Survey (TTS). Included are all Traffic Analysis
Zones (TAZ), the number of people who are employed full-time per TAZ,
the number of jobs per TAZ, the count of origin destination (OD) pairs
and trips by mode per origin TAZ, calculated car travel time from TAZ OD
centroid pairs, and associated spatial boundaries to link TAZ to the
Canadian Census. To illustrate how this information can be analysed to
understand patterns in commuting, we estimate a distance-decay curve
(i.e., impedance function) for the region. The value in \{TTS2016R\} is
that it is a growing open data product built on \texttt{R}
infrastructure that allows for the immediate access of home-to-work
commuting data alongside complimentary objects from different sources.
The package will continue expanding with additions by the authors and
the community at-large by requests in the future. \{TTS2016R\} can be
freely explored and downloaded in the associated
\href{https://github.com/soukhova/TTS2016R}{Github repository} where the
documentation and code involved in data creation (including this
manuscript), manipulation, and all open data products are detailed.
\end{abstract}

\keywords{Jobs; population; work; commute; travel time; impedance;
Greater Toronto and Hamilton Area; Ontario, Canada; R}

@@

This manuscript presents the open data product
\href{https://github.com/soukhova/TTS2016R}{\{TTS2016R\}}. Open data
products are the result of turning source data (open or otherwise) into
accessible information that adds value to the original inputs
\citep[see][]{Arribas2021open}. The product presented in this paper is a
\texttt{R} data package which currently consists of a fusion of objects
from a variety of sources: home-to-work flows sourced from the 2016
Transportation Tomorrow Survey (TTS)
\citep{data_management_group_tts_2018}, estimated travel times
(calculated using \{r5r\} \citep{Pereira2021r5r}), and boundary files
from the TTS \citep{datamanagementgroupSurveyBoundaryFiles2018} and from
the Canadian Census \citep{statisticscanadaBoundaryFiles20162017}.

What is a \texttt{R} data package? A \texttt{R} data package contains
code, data, and documentation in a standardised collection format that
can be installed by \texttt{R} users through a centralized software
repository such as CRAN (the Comprehensive R Archive Network) and
GitHub. \{TTS2016R\} is freely available on GitHub for all to install
and freely use in the spirit of open and reproducible research.
Currently and in more detail, \{TTS2016R\} includes full-time home-based
work-to-job origin destinations (OD) counts and mode-specific trip
numbers retrieved from the 2016 TTS, traffic analysis zone (TAZ)
boundaries, and municipality, planning, and census metropolitan area
boundaries for the Greater Golden Horse area (GGH) located in southern
Ontario, Canada. In addition, the package includes TAZ
centroid-to-centroid travel times by car, transit, cycling, and walking
mode computed using package \{r5r\} \citep{Pereira2021r5r}.

The aim of this paper is to walk readers through the data sets,
illustrate a use case (i.e., the calculation of an impedance function
that can be used to calculate accessibility to employment), and invite
others to experiment in its uses and applications. Though data from the
TTS is freely available to the public through the
\href{https://dmg.utoronto.ca/idrs/index}{TTS Data Retrieval System},
the raw data can be technically demanding, cumbersome to work with, and
requires multiple software to process. By pre-processing the data,
packaging it with complimentary data, and providing explicit
documentation in a \texttt{R} environment, \{TTS2016R\} offers a slice
of the TTS data that can be immediately used by \texttt{R} users to
analysis patterns of commuting to work in the region. Anticipate this
package to grow in the future: it currently provides an open
infrastructure for additional TTS or complimentary data sets to be
amended by the authors and the open-source community in the future by
request.

@@

This data is aggregated and available at the level of TAZ: TAZ are a
spatial unit of analysis typically used to estimate the number of trips
produced and attracted to each zone \citep{meyer_urban_2001}. They are
thus defined by transportation planners for a region based on
intra-similarity and inter-dissimilarity between land-use and population
demographics. Within the GGH boundaries, 3,764 TAZ are specified and
each TAZ is uniquely identified using the GTA06 Zoning System: the
survey boundary is discussed in the 2016 TTS methodology and defined by
the TTS \citep{data_management_group_tts_2018}. The TAZ range between
\(\ge\) 0.019 km\textsuperscript{2} in spatial area to a maximum of 879
km\textsuperscript{2} (median: 1.3 km\textsuperscript{2} and 3rd
quantile: 2.8 km\textsuperscript{2}).

@@

\caption{\label{fig:TTS-16-survey-area}TTS 2016 study area within the GGH in Ontario, Canada along with associated descriptive statistic of workers and jobs per TAZ, OD links (count of workers potentially interacting with their place of employment) by origin TAZ, and calculated OD car travel time (TT) per origin TAZ. 3,507 trips were not assigned TT as they are longer than 180 mins. Spatial boundary files are retrieved from the TTS which define the survey area (Data Management Group, 2018a): the 20 regions in the GGH are represented by black lines and labelled, the dark gray lines are planning boundaries.}\label{fig:TTS-16-survey-area}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/tts-workers-jobs-plot-1} \caption{\label{fig:tts-workers-jobs-plot}Number of workers (left) and jobs (right) in each TAZ retrieved from the 2016 TTS (Data Management Group, 2018b). Spatial boundary files are retrieved from the TAZ defined by the TTS (Data Management Group, 2018a).}\label{fig:tts-workers-jobs-plot}
\end{figure}

Figure \ref{fig:tts-workers-jobs-plot} presents the number of employed
people and associated jobs per TAZ. It can be observed that the spatial
distribution of jobs and workers is unequal, which is indicative of a
jobs -housing imbalance that can impact accessibility in a region
\citep{Levine1998rethinking}. It can also be seen that there is a higher
number of TAZ with no workers than zones with no jobs (i.e., 791 TAZ
with no workers : 396 TAZ with no jobs) and the mean of workers per TAZ
is higher than the mean of jobs. The number TAZ with an extreme number
of jobs at the highest and lowest percentiles is significantly higher
than the number of workers.

@@

Also included in \{TTS2016R\} are the estimated travel times between OD
as summarized in descriptive statistics table in Figure
\ref{fig:plot-tt-ttpertrip}; travel times are calculated using the
package \{r5r\}. \{r5r\} interfaces with the java-based R5 routing
engine developed separately by Conveyal
\citep{conveyalConveyalR5Routing2022}. The inputs to \{r5r\} for this
data package were the desired mode, a maximum travel time threshold of
180 minutes, the geo-coded origin destination pairs based on the
centroids of the TAZ, and the static Open Street Map road network of
Ontario (retrieved using Geofabrick
\citep{geofabrikOntarioOpenStreetMapGeofabrik2022}). A travel time
threshold of 180 minutes was selected since it captures almost all
potential OD interactions.

@@

\begin{figure}
\includegraphics[width=1\linewidth]{Manuscript-Data-Package_files/figure-latex/plot-tt-ttpertrip-1} \caption{\label{fig:plot-tt-ttpertrip}Calculated total worker travel time (left) and average worker travel time (right) for each TAZ in the 2016 TTS. Planning boundaries of Niagara and Waterloo (Data Management Group, 2018a), and the Toronto census metropolitan area (Statistics Canada, 2017) are drawn with purple, brown and blue borders, respectively. }\label{fig:plot-tt-ttpertrip}
\end{figure}
\newpage
